---
title: Process ALS point clouds to tree-level structure metrics
author: Johannes Heisig
always_allow_html: true
output:
  github_document:
    toc: true
---

```{r rglsetup, echo=FALSE, include=FALSE, eval=TRUE}
options(rgl.useNULL = TRUE)
knitr::opts_chunk$set(fig.width=8.5)
```

```{r message=FALSE, warning=FALSE}
# packages
library(sf)
library(stars)
library(lidR)
library(dplyr)
library(mapview)
library(tictoc)
library(pbmcapply)
library(readr)
library(purrr)
library(tidymodels)

# directories and functions
metrics_dir = "data/ALS/metrics"
als_dir = "data/ALS/als_clips"
fc_dir = "data/FuelCalc"
for (d in c(metrics_dir, als_dir, fc_dir)) if (!dir.exists(d)) dir.create(d, recursive = T)
source("R/00_lidar_metrics.R")

# species raster
species_codes = read.csv("data/Raster/tree_species_code.csv")
tree_spec = read_stars("data/Raster/species_10m.tif") |> 
  setNames("species") |> 
  mutate(species = factor(species, levels = species_codes$ID, 
                          labels = species_codes$name))
```

# Process tree-level metrics

## ALS data

Download an open LiDAR tile from the state of Northrhine-Westphalia.

```{r}
als_file = "data/ALS/ALS_demo_tile.laz"

if (!file.exists(als_file)){
  url = "https://www.opengeodata.nrw.de/produkte/geobasis/hm/3dm_l_las/3dm_l_las/3dm_32_387_5698_1_nw.laz"
  download.file(url, als_file)
}
```


Read ALS tile(s) into a catalog object. This demo uses just one tile, but you can provide multiple file names. Select only attributes of interest to save memory. In case of neighboring tiles, apply a buffer to prevent edge problems.

```{r}
ctg = readLAScatalog(als_file, select = "xyzc")
opt_chunk_size(ctg) = 100
opt_chunk_buffer(ctg) = 15
opt_select(ctg) = "xyzc"
summary(ctg)
```

## GEDI locations

Serve as plot locations.

```{r}
g = st_read("data/GEDI/GEDI_demo_subset.gpkg", quiet=T) |> 
  select(uniqueID) |> 
  st_transform(st_crs(ctg$geometry))

plot(ctg, map=F) 
plot(g["uniqueID"], col=1:nrow(g), add=T)
```

```{r eval=FALSE, include=FALSE}
plot(ctg, map=T) + mapview(g)
```

## Tree-level metrics

Process point clouds to tree-level structure metrics, including tree height, tree base height, and position.

1. clip at GEDI coordinate with 12.5 m radius
2. normalize height using a ground surface that is modeling useing a Kriging method (40 nearest neighbors)
3. segment individual trees using a region growing algorithm
4. derive tree level height metrics (max height and crown base height)

This routine is run in parallel to save time. It ouputs the segmented point cloud (`.las`) and the georeferenced metrics (`.gpkg`).

```{r echo=TRUE}
# takes ~ 30 min
metrics_list = pbmclapply(1:nrow(g), process_tlm2, gedi = g, .ctg = ctg, 
           alsdir = als_dir, mdir = metrics_dir, mc.cores = 4)
```

## Check example results

```{r, fig.height=8, fig.width=8}
sample_uID = 1145111
sample_metrics = st_read(paste0(metrics_dir, "/uID_", sample_uID, ".gpkg"), quiet = T)
sample_als = readLAS(paste0(als_dir, "/uID_", sample_uID, ".las"))

# map
par(bg = 'grey', fg = 'black')
plot(sample_metrics["treeID"], pch = 16, cex = sample_metrics$share*100)

# trees
plot3D::scatter3D(sample_als$X, sample_als$Y, sample_als$Z, 
                 colvar = sample_als$treeID, phi=30, pch = 16, 
                 cex=0.4, alpha = 0.6, col.panel = "black",
                 colkey = list(at = 1:max(sample_als$treeID),
                               side = 4, addlines = TRUE,
                               length = 0.9, width = 0.7,
                               labels = 1:max(sample_als$treeID),
                               col.axis = "white", col.clab = "white"))
```

```{r eval=FALSE, include=FALSE}
# map
mapview(sample_metrics, zcol = "Zmax")

# trees
plot(sample_als, color = "treeID", size = 4, axis =T)
rgl::rglwidget()
```

## Gather metrics

```{r}
files_metrics = list.files(metrics_dir, full.names = T)
files_als = list.files(als_dir, full.names = T)

ids_metrics = parse_number(basename(files_metrics))
ids_als = parse_number(basename(files_als))
identical(ids_als, ids_metrics)

read_with_id = function(path){
  id = parse_number(basename(path))
  m = st_read(path, quiet=T)
  m$uniqueID = id
  return(select(m, uniqueID, treeID, Z = Zmax, everything()))
}

metrics = map_dfr(files_metrics, read_with_id) |> 
  st_transform(st_crs(tree_spec))
```

## Add species label

Display tree locations on species raster and extract.

```{r fig.width=9, fig.height=8, message=FALSE, warning=FALSE}
metrics$species = st_extract(tree_spec, metrics) |> pull(1)
summary(metrics$species) # 69 species labels missing

plot(tree_spec, key.pos=1, key.length = 1, reset=F)
plot(metrics, pch = 21, bg = rainbow(10)[metrics$treeID], 
     col = "grey30", cex = 0.5, add=T)
```

## Handle missing species labels

Find dominant species per plot.

```{r}
dominant_species = metrics |> 
  st_drop_geometry() |> 
  group_by(uniqueID) |> 
  count(species) |> 
  summarise(Ndom = sum(n), dom_spec = species[which.max(n)], 
            perc_dom = n[which.max(n)]/Ndom, 
            Nna = sum(is.na(species)),
            perc_na = Nna/Ndom, .groups = "drop") 

summary(dominant_species[,2:6])
```

Insert dominant species label where missing.

```{r}
metrics_complete = inner_join(metrics, select(dominant_species, uniqueID, dom_spec), 
                              by = "uniqueID") |> 
  mutate(species = case_when(is.na(species) ~ dom_spec, TRUE ~ species)) |> 
  filter(sum(is.na(species)) == 0, .by = uniqueID)
```

# Prepare tree lists for Fuel Calc

## DBH and Crown Class

Predict diameter at breast height (DBH) with pre-trained models based on German national forest inventory data.

```{r}
DBHmodels = readRDS("data/ALS/DBH_models.rds")

for (i in 1:nrow(DBHmodels)){
  S = DBHmodels$species[i]
  metrics_complete$DBH[metrics_complete$species == S] = 
    predict(DBHmodels$fit_wf[[i]], 
            filter(metrics_complete, species == S) |> 
              select(Z)) |> 
    pull() |> round()
}
```

Infer Crown Class based on top position within canopy. Can be dominant (D), co-dominant (C), or suppressed (S).

```{r}
crcl = function(Z){
  if (length(Z) == 1){
    return(c(cl = as.factor("D")))
  } else{
  m = mean(Z, na.rm=T)
  s = sd(Z, na.rm=T)
  up = m + s
  lo = m - s 
  cut(Z, breaks = c(0, lo, up, 100), labels = c("S","C","D"))
  }
}

metrics_complete$CrCl = group_by(metrics_complete, uniqueID) |> 
  group_map(~crcl(.$Z)) |> unlist()
```

Fix unrealistic DBH and CBH.

```{r}
unrealistic = sum(metrics_complete$CBH > metrics_complete$Z*0.9 | metrics_complete$CBH < 2)
unrealistic / nrow(metrics_complete) # ~9 %
sum(metrics_complete$DBH < 70) / nrow(metrics_complete) # 0 %

metrics_complete = mutate(metrics_complete, 
                          CBH = ifelse(CBH > Z*0.9 | CBH < 2, Z/2, CBH)) |>       
  filter(DBH > 70)
```
 
```{r}
summary(metrics_complete)
par(mfrow=c(1,3))
hist(metrics_complete$Z, main = "Z", xlab = "m")      # canopy height
hist(metrics_complete$CBH, main = "CBH", xlab = "m")  # canopy base height
hist(metrics_complete$DBH, main = "DBH", xlab = "mm") # diameter breast height
t_spec = table(metrics_complete$species)
t_spec[t_spec > 0]
```

## Export to FuelCalc

1. `.tre` file containing tree lists

```{r}
export_treelist = function(x){
  
  t = data.frame(
    PlotId = paste0("ID_", x$uniqueID), 
    MonStatus = "Inventory", MonOrder = 1,
    TreeSpecies = as.character(x$species), 
    TreeExpansionFactor = 8.25,                    # Trees Per Acre: 4047 / (12.5*12.5*pi)  = ~ 8.25   
    Diameter = round(x$DBH/10 / 2.54),             # diameter mm in inches
    TreeHeight = round(x$Z * 3.2808399),
    CrownBaseHeight = round(x$CBH * 3.2808399),    # heights in feet
    TreeStatus = "H",
    CrownClass = x$CrCl, CrownRatio = "", CharHeight = "",
    CrownScorchPercent = "", CrownScorchHeight = "", CKR = "",
    BeetleDamage = "No", EquationType = "CRNSCH", FlHt.ScHt = 0,
    FS = "F", Severity = "" )
  
  t = mutate(t, TreeSpecies = case_when( 
    TreeSpecies == "Birch" ~ "BIRCH",      # = Yellow Birch
    TreeSpecies == "Beech" ~ "BEECH",      # = Tan Oak
    TreeSpecies == "Douglas_fir" ~ "DFIR", # = Douglas Fir
    TreeSpecies == "Oak" ~ "OAK",          # = Northern Red Oak
    TreeSpecies == "Alder" ~ "ALDER",      # = Red Alder
    TreeSpecies == "Spruce" ~ "SPRUCE",    # = Engelman Spruce
    TreeSpecies == "Pine" ~ "PINE",        # = Lodgepole Pine
    TreeSpecies == "Larch" ~ "LARCH",      # = Western Larch
    TreeSpecies == "Fir" ~ "FIR",          # = Grand Fir
    TreeSpecies == "ODH" ~ "ODH",          # = Quaking Aspen
    TreeSpecies == "ODL" ~ "ODL",          # = Quaking Aspen
    TRUE ~ TreeSpecies))
  return(t)
}

tl = export_treelist(metrics_complete)
head(tl, 2)
write.csv(tl, file.path(fc_dir,"input_treelist.tre"), row.names = F, quote = F)
```

2. `.ffi` file containing plot info

```{r}
export_plotlist = function(x){
  ffi = read.csv(file.path(fc_dir, "test_file.FFI"))
  fh = data.frame(matrix(ncol=ncol(ffi), nrow = length(unique(x$uniqueID))))
  names(fh) = names(ffi)
  fh[] = 1
  fh$PlotId = paste0("ID_", unique(x$uniqueID))
  fh$MonStatus = "Inventory"
  fh$EmisCWDRSC = fh$EmisDuffRSC = fh$EmisSTFS = ""
  return(fh)
}

pl = export_plotlist(metrics_complete)
head(pl, 2)

write.csv(pl, file.path(fc_dir,"input_treelist.ffi"), row.names = F, quote = F)
```


